A Formal Verification Framework for Smart Contracts Based on Symbolic Execution

### 2024-01-17 papers/static-checker/ecoop2024/ecoop2024.tex
intro & abstract

\documentclass[a4paper,USenglish,cleveref, autoref, thm-restate]{lipics-v2021}

@@

\ccsdesc[100]{\textcolor{red}{Replace ccsdesc macro with valid one}} %TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm

@@

%TODO mandatory: add short abstract of the document
\begin{abstract}
Formal verification is a requirement for all but the simplest smart
contracts on the blockchain. We developed SCV, a tool that is geared towards contracts
implemented in Michelson, the smart contract language of the Tezos
blockchain. It relies on symbolic execution and an SMT solver to
derive verification conditions that can be checked against the user's
specification. Specifications are written in a domain-specific
language tailored to Michelson's execution model.

We report the foundations of the design and implementation of SCV and
conduct two case studies that demonstrate the expressiveness of our tool.
\end{abstract}

@@

Once a smart contract is published on a blockchain, it cannot be
changed any more. Unfortunately, there are plenty of
cases~\cite{dao,wallethack} where the blockchain 
community identified vulnerabilities in smart contracts despite
extensive testing before deployment.
In addition, the novel execution model of smart contract
languages has led to unforeseen errors due to a lack of familiarity
with their intricacies. These observations along with the contract's
immutability underscore the need to ensure its correctness before deployment.

@@

In this work, we present the design of a domain-specific language for
specifying smart contract properties and demonstrate the integration
of symbolic execution for Michelson with an SMT solver for automatic
formal verification of these properties.

\begin{lstlisting}[float=tp,captionpos=b,caption={Auction contract specification (excerpt)},label={lst:auction-fragment}]
mcontract auction = spec 
  storage := Pair (auction_open: bool) 
                  Pair (highest_bidder: address)
                       (contract_owner: address))
  entrypoint %bid
    code := {...}
    parameter := Unit 
    pre-condition := auction_open 
       && Amount > pre_Balance 
       && get_contract (unit, highest_bidder) = 
          Some (c: contract unit)
    post-condition := auction_open
       && post_highest_bidder = Sender 
       && post_contract_owner = contract_owner 
       && post_Balance = Amount 
       && transfer_token (Unit, Old_Balance, c) ;
\end{lstlisting}
As an example, Listing~\ref{lst:auction-fragment} contains an excerpt
of the specification of an auction contract. The \texttt{storage}
phrase describes the internal state (storage) of the contract
consisting of three values named \lstinline|auction_open|,
\lstinline|highest_bidder|, and \lstinline|contract_owner|.
The actual specification consists of a precondition and a
postcondition with the usual partial correctness semantics. The
postcondition refers to components of the final storage by prefixing
\lstinline|post_| to their name. The variables
\lstinline|Amount|, and \lstinline|Sender| correspond to the values
returned by the like-named Michelson instructions that return, the amount transferred with the call, and the address of the caller (sender). The variables \lstinline|pre_Balance| and \lstinline|post_Balance| represent the previous balance of the contract right before and after the call, respectively. The current balance of the contract returned by the instruction \BALANCE\ includes \lstinline|pre_Balance| and the \lstinline|Amount|.

@@

@@

Symbolic execution has emerged as a powerful technique for systematically exploring program paths and identifying potential vulnerabilities in smart contracts. P. Tsankov et al. introduced SECUTIFY \cite{securify}, a tool that utilizes symbolic execution to perform practical security analysis on Ethereum smart contracts. It targets common vulnerability security patterns specified in a designated domain-specific language. SECUTIFY symbolically encodes the dependence graph of the contract in stratified Datalog to extract semantic information from the code. After obtaining semantic facts, it checks whether the security patterns hold or not. Similarly, Manticore \cite{manticore} and KEVM \cite{kevm} use symbolic execution to analyze Ethereum smart contracts. KEVM is an executable formal specification built with the K Framework for the Ethereum virtual machine's bytecode (EVM), a stack-based and low-level smart contract language for the Ethereum blockchain. Since tokens can hold a significant amount of value, they are often targeted for attacks. Therefore, several tools \cite{kevm, park} conduct case studies for the implementations of token standards.

Several approaches use existing formal verification frameworks to ensure the correctness and security of smart contracts. S. Amani et al. \cite{isabelle} proposed the formal verification of Ethereum smart contracts in Isabelle/HOL. Hirai \cite{hirai} formalizes the EVM using Lem, a language to specify semantic definitions. The formal verification of smart contracts is achieved using the Isabelle proof assistant. Bernardo et al. \cite{micho} implemented Mi-Cho-Coq, a formalization of the semantics of Michelson using the Coq proof assistant. They also verified several Michelson contracts.

>>>>

\documentclass[a4paper,USenglish,cleveref, autoref,anonymous]{lipics-v2021}

@@

\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10011007.10010940.10010992.10010998.10011000</concept_id>
<concept_desc>Software and its engineering~Automated static analysis</concept_desc>
<concept_significance>500</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~Automated static analysis}

@@

%TODO mandatory: add short abstract of the document
\begin{abstract}
Formal verification is a requirement for all smart
contracts on the blockchain. We developed SCV, a verification tool
that is geared towards contracts implemented in Michelson, the smart
contract language of the Tezos blockchain. It relies on symbolic execution to
derive verification conditions that can be checked automatically against the user's
specification. An SMT solver helps prune states during symbolic
execution and check the verification conditions against the
specification. Specifications are written in a domain-specific
language tailored to Michelson's execution model.

We report the foundations of the design and implementation of SCV and
conduct two case studies on real-life contracts that demonstrate the
expressiveness of our tool.
\end{abstract}

@@

A blockchain is a unique execution environment rooted in
irreversible transactions. Smart contracts, which are programs
designed for the blockchain, are also permanently stored within
it. Consequently, once a smart contract is deployed on a blockchain,
it cannot be changed anymore. Regrettably, numerous
instances~\cite{dao,wallethack} have been documented where
vulnerabilities in smart contracts were 
identified by the blockchain community, resulting in substantial
financial losses. Specifically, the distinctive execution model of
smart contract languages has given rise to unforeseen errors due to a
lack of familiarity with their complexities. These observations,
combined with the immutability of contracts, underscore the need
to ensure their correctness prior to deployment.

@@

Our starting point is the design of a domain-specific language for
specifying smart contract properties.
\begin{lstlisting}[float=tp,captionpos=b,caption={Auction contract specification (excerpt)},label={lst:auction-fragment}]
mcontract auction = spec 
  storage := Pair (auction_open: bool) 
                  Pair (highest_bidder: address)
                       (contract_owner: address))
  entrypoint %bid
    code := {...}
    parameter := Unit 
    pre-condition := auction_open 
       && Amount > pre_Balance 
       && get_contract (unit, highest_bidder) = 
          Some (c: contract unit)
    post-condition := auction_open
       && post_highest_bidder = Sender 
       && post_contract_owner = contract_owner 
       && post_Balance = Amount 
       && transfer_token (Unit, Old_Balance, c) ;
\end{lstlisting}
As an example, Listing~\ref{lst:auction-fragment} contains an excerpt
of the specification of an auction contract. The \texttt{storage}
phrase describes the internal state (storage) of the contract
consisting of three values named \lstinline|auction_open|,
\lstinline|highest_bidder|, and \lstinline|contract_owner|.
The contract has several entrypoints that may be invoked individually.

There is a specification for each entrypoint. The example shows the
one for  \lstinline|bid|, to place a bid on the auction. Each
specification consists of a precondition and a 
postcondition with the usual partial correctness semantics. The
postcondition refers to components of the final storage by prefixing
\lstinline|post_| to their name. The variables
\lstinline|Amount| and \lstinline|Sender| correspond to the values
returned by the like-named Michelson instructions that return the
amount transferred with the call and the address of the caller
(sender).  The variables \lstinline|pre_Balance| and
\lstinline|post_Balance| stand for the balance of the contract right
before and after the call, respectively. The value returned by the
Michelson instruction \BALANCE\ includes 
\lstinline|pre_Balance| and the \lstinline|Amount|.
This contract is our running example in
Section~\ref{sec:more-than-one}, where further explanations may be found.

@@

While there is already a considerable body of work that applies
symbolic execution to the verification of smart contracts
\cite{securify, manticore,kevm,park} (see
Section~\ref{sec:related-work} for details), most of them are geared
towards Ethereum, so that standard techniques for symbolic execution
of imperative programs are applicable. Such groundwork does not exist
for Michelson, which is a functional language running on a stack
machine (see Section~\ref{sec:background}) with a sophisticated
instruction set that includes high-level loops. For that reason, we had to
develop symbolic execution for Michelson from ground up.

Besides the Mi-Cho-Coq framework that relies on interactive theorem
proving \cite{micho}, HELMHOLTZ \cite{helmholtz} is another automatic
verifier for Michelson. However, the approach of HELMHOLTZ is significantly
different from ours. They define a refinement type system to keep
track of the properties of values in the storage and on the
stack, so that verifying a program corresponds to typechecking against
a specification type. Their type checker employs Z3 to discharge proof obligations.

@@

Symbolic execution has emerged as a powerful technique for
systematically exploring program paths and identifying potential
vulnerabilities in smart contracts. P. Tsankov et al. introduced
SECURIFY \cite{securify}, a tool that utilizes symbolic execution to
perform practical security analysis on Ethereum smart contracts. It
targets common vulnerability security patterns specified in a
designated domain-specific language. SECURIFY symbolically encodes the
dependence graph of the contract in stratified Datalog to extract
semantic information from the code. After obtaining semantic facts, it
checks whether the security patterns hold or not. Similarly, Manticore
\cite{manticore} and KEVM \cite{kevm} use symbolic execution to
analyze Ethereum smart contracts. KEVM is an executable formal
specification built with the K Framework for the Ethereum virtual
machine's bytecode (EVM), a stack-based and low-level smart contract
language for the Ethereum blockchain. Since tokens can hold a
significant amount of value, they are often targeted for
attacks. Therefore, several tools \cite{kevm,park} conduct case
studies for the implementations of token standards.

Several approaches use existing formal verification frameworks to
ensure the correctness and security of smart contracts. S. Amani et
al. \cite{isabelle} proposed the formal verification of Ethereum smart
contracts in Isabelle/HOL. Hirai \cite{hirai} formalizes the EVM using
Lem, a language to specify semantic definitions. The formal
verification of smart contracts is achieved using the Isabelle proof
assistant. Bernardo et al.~\cite{micho} implemented Mi-Cho-Coq, a
formalization of the semantics of Michelson using the Coq proof
assistant. They also verified several Michelson contracts.
